---
title: "bulkMash_simu_replicate10"
author: "DongyueXie"
date: "2020-06-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In a previous [simulation](bulkMash_simu.html), the number of replicates in each condition is 3, in which only DESeq2 controls false discoveries well. Here, the number of replicates in each condition is increased to 10.


```{r}
library(mashr)
fdp = function(dis.idx, true.idx){
  if(length(dis.idx)==0){
    0
  }else{
    1-mean(dis.idx%in%true.idx)
  }
}


auc = function(pred,true.label){
  auc=pROC::roc(response = true.label, predictor = pred)
  auc$auc
}

powr = function(dis.idx, true.idx){
  if(length(dis.idx)==0){
    0
  }else{
    sum(dis.idx%in%true.idx)/length(true.idx)
  }
}
```

# NULL

## Compare to mean

```{r}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_null_replicate10.RData')
nsimu=10

methods = names(null_simu)
method = methods[c(1,2,3,4)]

fdp_all = matrix(nrow=nsimu,ncol=length(method))
n_fdis = matrix(nrow=nsimu,ncol=length(method))
for(i in 1:nsimu){
  
  for(j in 1:length(method)){
    
    m = which(names(null_simu)==method[j])
    
    fdp_all[i,j] = fdp(get_significant_results(null_simu[[m]][[i]]$mash),0) 
    n_fdis[i,j] = length(get_significant_results(null_simu[[m]][[i]]$mash))
    
  }

}
colnames(fdp_all) = method
colnames(n_fdis) = method
rownames(n_fdis) = 1:nsimu
knitr::kable(n_fdis,caption='number of false discoveries')

```

Let's look at plots of $z$ scores of the first simulation since voom-limma has a number of false discoveries. compare limma, voom-limma and deseq

### limma z score

```{r,fig.width=9,fig.height=9}
Bhat = null_simu$limma[[1]]$mash$input$Bhat
Shat = null_simu$limma[[1]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```

### voom-limma z score

```{r,fig.width=9,fig.height=9}
Bhat = null_simu$voom_limma[[1]]$mash$input$Bhat
Shat = null_simu$voom_limma[[1]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```


### rle-voom-limma z score

```{r,fig.width=9,fig.height=9}
Bhat = null_simu$rle_voom_limma[[1]]$mash$input$Bhat
Shat = null_simu$rle_voom_limma[[1]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```

### deseq z score

```{r,fig.width=9,fig.height=9}
Bhat = null_simu$deseq[[1]]$mash$input$Bhat
Shat = null_simu$deseq[[1]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.6))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```





# Signal

## Compare to mean

```{r}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_signal_mean_ref_replicate10.RData')
nsimu=10
methods = names(signal_simu_mean)
method = methods[c(1:4)]

fdp_signal = matrix(nrow=nsimu,ncol=length(method))
auc_signal = matrix(nrow=nsimu,ncol=length(method))
power_signal = matrix(nrow=nsimu,ncol=length(method))



for(i in 1:nsimu){
  
  which_null = 1*(rowSums(signal_simu_mean$non_null_matrix[[i]])==0)
  non_null_idx = which(rowSums(signal_simu_mean$non_null_matrix[[i]])!=0)
  

  
  for(j in 1:length(method)){
    
    m = which(names(signal_simu_mean)==method[j])
    
    fdp_signal[i,j] = fdp(get_significant_results(signal_simu_mean[[m]][[i]]$mash),non_null_idx) 
    auc_signal[i,j] = auc(apply(signal_simu_mean[[m]][[i]]$mash$result$lfsr,1,min),which_null)
    power_signal[i,j] = powr(get_significant_results(signal_simu_mean[[m]][[i]]$mash),non_null_idx) 
    
  }
  
  
}
colnames(fdp_signal) = method
colnames(auc_signal) = method
rownames(auc_signal) = 1:nsimu
colnames(power_signal) = method


knitr::kable(sort(round(apply(fdp_signal,2,mean),2)),col.names = 'lfsr level 0.05')
knitr::kable(sort(round(apply(auc_signal,2,mean),2),decreasing = TRUE),col.names = 'auc')
knitr::kable(sort(round(apply(power_signal,2,mean),2),decreasing = TRUE),col.names = 'power at lfsr level 0.05')
```

```{r,fig.width=15,fig.height=5}
par(mfrow=c(1,3))
boxplot(fdp_signal,main='fdp')
boxplot(auc_signal,main='auc')
boxplot(power_signal,main='power')
```


## Compare to control



```{r}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_signal_control_ref_replicate10.RData')
nsimu=10
methods = names(signal_simu_control)
method = methods[c(1,2,3,4)]

fdp_signal = matrix(nrow=nsimu,ncol=length(method))
auc_signal = matrix(nrow=nsimu,ncol=length(method))
power_signal = matrix(nrow=nsimu,ncol=length(method))

for(i in 1:nsimu){
  
  which_null = 1*(rowSums(signal_simu_control$non_null_matrix[[i]])==0)
  non_null_idx = which(rowSums(signal_simu_control$non_null_matrix[[i]])!=0)
  
  for(j in 1:length(method)){
    
    m = which(names(signal_simu_control)==method[j])
    
    fdp_signal[i,j] = fdp(get_significant_results(signal_simu_control[[m]][[i]]$mash),non_null_idx) 
    auc_signal[i,j] = auc(apply(signal_simu_control[[m]][[i]]$mash$result$lfsr,1,min),which_null)
    power_signal[i,j] = powr(get_significant_results(signal_simu_control[[m]][[i]]$mash),non_null_idx) 
    
  }
  
  
}
colnames(fdp_signal) = method
colnames(auc_signal) = method
rownames(auc_signal) = 1:nsimu
colnames(power_signal) = method


knitr::kable(sort(round(apply(fdp_signal,2,mean),2)),col.names = 'lfsr level 0.05')
knitr::kable(sort(round(apply(auc_signal,2,mean),2),decreasing = TRUE),col.names = 'auc')
knitr::kable(sort(round(apply(power_signal,2,mean),2),decreasing = TRUE),col.names = 'power')
```

```{r,fig.width=15,fig.height=5}
par(mfrow=c(1,3))
boxplot(fdp_signal,main='fdp')
boxplot(auc_signal,main='auc')
boxplot(power_signal,main='power')
```

## other methods based on p-value

```{r}
methods = names(signal_simu_control)
method = methods[c(5,6,7)]

fdp_signal = matrix(nrow=nsimu,ncol=length(method))
auc_signal = matrix(nrow=nsimu,ncol=length(method))
power_signal = matrix(nrow=nsimu,ncol=length(method))

for(i in 1:nsimu){
  
  which_null = 1*(rowSums(signal_simu_control$non_null_matrix[[i]])==0)
  non_null_idx = which(rowSums(signal_simu_control$non_null_matrix[[i]])!=0)
  
  for(j in 1:length(method)){
    
    m = which(names(signal_simu_control)==method[j])
    
    fdp_signal[i,j] = fdp((signal_simu_control[[m]][[i]]$sig_idx),non_null_idx) 
    auc_signal[i,j] = auc(p.adjust(signal_simu_control[[m]][[i]]$p_val,method='BH'),which_null)
    power_signal[i,j] = powr(signal_simu_control[[m]][[i]]$sig_idx,non_null_idx) 
    
  }
  
  
}
colnames(fdp_signal) = method
colnames(auc_signal) = method
rownames(auc_signal) = 1:nsimu
colnames(power_signal) = method


knitr::kable(sort(round(apply(fdp_signal,2,mean),2)),col.names = 'fdr level 0.05')
knitr::kable(sort(round(apply(auc_signal,2,mean),2),decreasing = TRUE),col.names = 'auc')
knitr::kable(sort(round(apply(power_signal,2,mean),2),decreasing = TRUE),col.names = 'power')
```

```{r,fig.width=15,fig.height=5}
par(mfrow=c(1,3))
boxplot(fdp_signal,main='fdp')
boxplot(auc_signal,main='auc')
boxplot(power_signal,main='power')
```

