---
title: "bulkMash_simu"
author: "DongyueXie"
date: "2020-06-19"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction

methods: 

1. voom-limma
2. RLE-voom-limma: apply RLE normalization first 
3. RLE-voom-limma_V: apply RLE normalization first and plug covariance matrix into mash
4. deseq

5. anova: simple anova applied to log(Y+0.5) after RLE normalziation
6. kw: Kruskal-Wallis Rank Sum Test
7. edgeR: likelihood ratio test


Design: 

1. NUll: each simulation, draw 2000 genes and 30 samples, randomly assign 30 samples to 10 groups with 3 replicates in each group.  

2. Signal: generate thinning coefficients from $N(0,0.8^2)$, randomly select $0.1$ genes to be non-NULL and for non-null genes, $3$ groups are selected to be non-null. Apply thinning to null matrix. 

Evaluation:

1. false discovery rate; 2. area under curve; 3. power.

Is there a ROC-like plot for fdr vs tpr?




```{r}
library(mashr)
source('code/pln_simu_summary.R')
```

# Three replicates

```{r,results='asis'}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_3rep.RData')
summary_pln_simu(simu_all)
```




Let's look at plots of $z$ scores since voom-limma has a number of false discoveries. Compare rle-voom-limma and deseq



```{r,fig.width=9,fig.height=9,eval=F,include=F}

### limma z score

Bhat = simu_all$null_simu_mean$limma[[2]]$mash$input$Bhat
Shat = null_simu$limma[[2]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```



```{r,fig.width=9,fig.height=9,eval=F,include=F}

### voom-limma z score

Bhat = null_simu$voom_limma[[2]]$mash$input$Bhat
Shat = null_simu$voom_limma[[2]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```


### rle-voom-limma z score

```{r,fig.width=9,fig.height=9}
Bhat = simu_all$null_simu_mean$rle_voom_limma[[9]]$mash$input$Bhat
Shat = simu_all$null_simu_mean$rle_voom_limma[[9]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.45))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```

### deseq z score

```{r,fig.width=9,fig.height=9}
Bhat = simu_all$null_simu_mean$deseq[[9]]$mash$input$Bhat
Shat = simu_all$null_simu_mean$deseq[[9]]$mash$input$Shat

data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')

titles = colnames(data.L$Bhat)

par(mfrow=c(3,3))
for(i in 1:length(titles)){
  z = data.L$Bhat[,i]/data.L$Shat[,i]
  hist(z,breaks = 30,xlab='z', main=titles[i],freq = FALSE,ylim = c(0,0.6))
  
  xfit<-seq(min(z),max(z),length=40)
  yfit<-dnorm(xfit,mean=0,sd=1)
  lines(xfit, yfit, col="red", lwd=2)
}

```


## Look at some specific false discoveies of voom-limma

```{r}
n_sig_cond = as.numeric(get_n_significant_conditions(simu_all$null_simu_mean$rle_voom_limma[[9]]$mash))
par(mfrow=c(1,1))
hist(n_sig_cond,breaks=100)
```

either none or most of conditions are significant....

```{r}
Bhat = simu_all$null_simu_mean$rle_voom_limma[[9]]$mash$input$Bhat
Shat = simu_all$null_simu_mean$rle_voom_limma[[9]]$mash$input$Shat
data = mash_set_data(Bhat,Shat)
data.L = mash_update_data(data,ref='mean')
data.L$Bhat[1484,]
data.L$Shat[1484,]
simu_all$null_simu_mean$rle_voom_limma[[9]]$mash$result$lfsr[1484,]
```

The results are weried. Look at large Shat compared to Bhat but small lfsr.

Let's run this specific see what happened.

```{r}
U.c = cov_canonical(data.L)
m.1by1 = mash_1by1(data.L)
strong = get_significant_results(m.1by1)
length(strong)
U.pca = cov_pca(data.L,2,subset=strong)
U.ed = cov_ed(data.L, U.pca, subset=strong)
m = mash(data.L, c(U.c,U.ed), algorithm.version = 'Rcpp',verbose = F)
length(get_significant_results(m))
m$loglik
```

So ash 1by1 identifies 69 significant genes while mash finds 243. How about just use canonical matrix?

```{r}
m = mash(data.L, c(U.c), algorithm.version = 'Rcpp',verbose = F)
length(get_significant_results(m))
m$loglik
```

Only 12! The prior covariance matrix has such huge influence on the results!!






# Ten replicates


```{r,results='asis'}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_10rep.RData')
summary_pln_simu(simu_all)
```









```{r,eval=F,include=F}

## other methods based on p-value

methods = names(signal_simu_control)
method = methods[c(5,6,7)]

fdp_signal = matrix(nrow=nsimu,ncol=length(method))
auc_signal = matrix(nrow=nsimu,ncol=length(method))
power_signal = matrix(nrow=nsimu,ncol=length(method))

for(i in 1:nsimu){
  
  which_null = 1*(rowSums(signal_simu_control$non_null_matrix[[i]])==0)
  non_null_idx = which(rowSums(signal_simu_control$non_null_matrix[[i]])!=0)
  
  for(j in 1:length(method)){
    
    m = which(names(signal_simu_control)==method[j])
    
    fdp_signal[i,j] = fdp((signal_simu_control[[m]][[i]]$sig_idx),non_null_idx) 
    auc_signal[i,j] = auc(p.adjust(signal_simu_control[[m]][[i]]$p_val,method='BH'),which_null)
    power_signal[i,j] = powr(signal_simu_control[[m]][[i]]$sig_idx,non_null_idx) 
    
  }
  
  
}
colnames(fdp_signal) = method
colnames(auc_signal) = method
rownames(auc_signal) = 1:nsimu
colnames(power_signal) = method


knitr::kable(sort(round(apply(fdp_signal,2,mean),2)),col.names = 'fdr level 0.05')
knitr::kable(sort(round(apply(auc_signal,2,mean),2),decreasing = TRUE),col.names = 'auc')
knitr::kable(sort(round(apply(power_signal,2,mean),2),decreasing = TRUE),col.names = 'power')


par(mfrow=c(1,3))
boxplot(fdp_signal,main='fdp')
boxplot(auc_signal,main='auc')
boxplot(power_signal,main='power')
```





















```{r,eval=F,include=F}
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_null.RData')
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_signal_mean_ref.RData')
load('/scratch/midway2/dyxie/sc-cytokine/output/bulkMash_cytokine_SI_signal_control_ref.RData')

simu_all = list(signal_simu_control=signal_simu_control,signal_simu_mean=signal_simu_mean)

nsimu=18

methods = names(null_simu)
method = methods[c(1,2,3,4)]

fdp_all = matrix(nrow=nsimu,ncol=length(method))
n_fdis = matrix(nrow=nsimu,ncol=length(method))
for(i in 1:nsimu){
  
  for(j in 1:length(method)){
    
    m = which(names(null_simu)==method[j])
    
    fdp_all[i,j] = fdp(get_significant_results(null_simu[[m]][[i]]$mash),0) 
    n_fdis[i,j] = length(get_significant_results(null_simu[[m]][[i]]$mash))
    
  }

}
colnames(fdp_all) = method
colnames(n_fdis) = method
rownames(n_fdis) = 1:nsimu
knitr::kable(n_fdis,caption='number of false discoveries')

```
