---
title: "compare  linear model and sample mean/var based methods"
author: "DongyueXie"
date: "2020-05-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Two methods: use limma / sample mean&var obtaining Bhat and Shat then feed into mashr-no-baseline. These two method give exactly the same Bhat but different Shat.

I want to compare those false discoveries from using sample mean&var. 

To do so, I filtered out genes that appear in less than 1000 cells out of total 13000+ cells. The main reason is to make model fitting faster, and make data cleaner. 

For each false discovery, I compared lfsr, Shat and distribution of z-scores from mashr_update_data. In the plot, lm stands for limma method and sample stands for sample mean&var.


```{r}
library(mashr)
library(ggplot2)
library(plyr)
load('output/compare_lm_sample.RData')

out = compare_lm_sample$out_lm
out.s = compare_lm_sample$out_sam
mash_data = mash_set_data(out$betahat,out$sebetahat)
mash_data.L = mash_update_data(mash_data,ref='mean')

mash_data.s = mash_set_data(out.s$Bhat,out.s$Shat)
mash_data.L.s = mash_update_data(mash_data.s,ref='mean')

get_significant_results(compare_lm_sample$mash_lm)
get_significant_results(compare_lm_sample$mash_sam)

```



```{r}

summary_mm = function(idx){
  #First look at lfsr. Big difference!
  dat = cbind(as.numeric(compare_lm_sample$mash_lm$result$lfsr[idx,]),
            as.numeric(compare_lm_sample$mash_sam$result$lfsr[idx,]))
  colnames(dat) = c('lm','sample')
  boxplot(dat,ylab='lfsr')
  
  #Look at corresponding S_hat: similar median but difference variance.
  dat = cbind(as.numeric(compare_lm_sample$out_lm$sebetahat[idx,]),
              as.numeric(compare_lm_sample$out_sam$Shat[idx,]))
  colnames(dat) = c('lm','sample')
  boxplot(dat,ylab = 'S_hat')
  
  #Distribution of z-statistics after mash_update_data
  
z1 = mash_data.L$Bhat[idx,]/mash_data.L$Shat[idx,]
z2 = mash_data.L.s$Bhat[idx,]/mash_data.L.s$Shat[idx,]
dat = data.frame(z = c(z1, z2), method = rep(c("lm", "sample"),each=49))

mu = ddply(dat, "method", summarise, grp.mean=mean(z))
ggplot(dat, aes(x = z, fill = method,color=method)) + 
  geom_histogram(alpha = 0.5,bins=50,position="identity") +
  geom_vline(data=mu, aes(xintercept=grp.mean, color=method),
             linetype="dashed") +
  theme(legend.position="top")
}
```

Let's look at some genes that are identified by mashr using sample mean and variance.

```{r}
summary_mm(2121)
```

The difference of lfsr are huge. $\hat s$ are of similar scale but the ones from sample estiamtion is more dispersed.  The plot of z-statistics shows that there is an extreme z-score of sample method. 


```{r}
summary_mm(1198)
```
      
Again, z-statistics from sample method has heavier tails.

```{r}
summary_mm(2161)
```

Let's take a look at this extreme z-score.

```{r}
as.numeric(mash_data.L.s$Bhat[2161,17])
as.numeric(mash_data.L.s$Shat[2161,17])
as.numeric(mash_data.L$Shat[2161,17])
```

The Bhat is -0.047, and Shat from sample is 0.0089 while 0.014 from limma. 

## Summary

1. Z-scores from sample method could have some extreme ones, especially negative ones. This is due to small Shat.
The conseqeunce is false postives.

2. On plausibility of multiple linear regression model: 

```{r}
X = model.matrix(~as.factor(compare_lm_sample$group_idx))
X.lm = cbind(X[,1]-rowSums(X[,-1,drop=FALSE]),X[,-1])
cov_of_interest = 1:ncol(X)

par(mfrow=c(2,2))

for(i in 1:4){
  lm1 = lm(y~.-1,data.frame(y=compare_lm_sample$Y[,i],x=X.lm))
  plot(fitted(lm1),resid(lm1),xlab='fitted',ylab='residual')
}
```
