---
title: "mash randomeffect extended"
author: "Dongyue Xie"
date: "2020-10-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Extend universal sigma2 [examples](mash_randomeffect.html).

We generate $X$ from a mixture of 2 multivariate normal distributions, $x_i\sim \sum_k\pi_k N(0,U_k+S_i)$, with the following covariance matrices: present (and identical) in first two conditions, present (and identical) in last three conditions, each with 500 samples(so $\pi_1=\pi_2=0.5$).

Try varying $\sigma^2$ and see how it effects the results.

## small error sd

$S_i=diag(0.1^2,...,0.1^2)$.

```{r}
source('code/myED.R')
source('code/mash_randomeffect.R')
```

```{r}
sigma2_list = c(0,1e-5,1e-4,1e-3,1e-2,1e-1,1)
name = paste('sigma2=',sigma2_list,sep='')
set.seed(12345)
simdata = simple_sims0(500,err_sd = 0.1)
data = mash_set_data(simdata$Bhat,simdata$Shat)
strong = 1:nrow(simdata$B)
U.c    = cov_canonical(data)
U.pca  = cov_pca(data,5,strong)
```

```{r}
# myed.all = list()
# myed.mash.all = list()
# sigma2_list = c(0,1e-5,1e-4,1e-3,1e-2,1e-1,1)
# name = c()
# for(isig in 1:length(sigma2_list)){
#   myed.universal = myED_wrapper(data,U.pca,sigma2_list[isig],fix_sigma2 = TRUE,maxiter=100,tol=0.01)
#   myed.all[[isig]] = myed.universal
#   U.myed.universal  = myMash_data(myed.universal,data)$Ulist
#   m.c.myed.universal = mash(data, c(U.c,U.myed.universal),verbose = F)
#   myed.mash.all[[isig]] = m.c.myed.universal
#   name[isig] = paste('sigma2=',sigma2_list[isig],sep='')
# }
# save(myed.all,file='output/myed_all_01.RData')
# save(myed.mash.all,file = "output/myed_mash_all_01.RData")
load('output/myed_mash_all_01.RData')
out = summary_out(simdata$B,myed.mash.all)
knitr::kable(out$find_cond,col.names = name,caption = 'On finding conditions',digits = 5)
```

## larger error sd

Try $S_i=diag(0.5^2,...,0.5^2)$

```{r}
set.seed(12345)
simdata = simple_sims0(500,err_sd = 0.5)
data = mash_set_data(simdata$Bhat,simdata$Shat)
strong = 1:nrow(simdata$B)
U.c    = cov_canonical(data)
U.pca  = cov_pca(data,5,strong)
```

```{r}
# myed.all = list()
# myed.mash.all = list()
# sigma2_list = c(0,1e-5,1e-4,1e-3,1e-2,1e-1,1)
# name = c()
# for(isig in 1:length(sigma2_list)){
#   myed.universal = myED_wrapper(data,U.pca,sigma2_list[isig],fix_sigma2 = TRUE,maxiter=100,tol=0.01)
#   myed.all[[isig]] = myed.universal
#   U.myed.universal  = myMash_data(myed.universal,data)$Ulist
#   m.c.myed.universal = mash(data, c(U.c,U.myed.universal),verbose = F)
#   myed.mash.all[[isig]] = m.c.myed.universal
#   name[isig] = paste('sigma2=',sigma2_list[isig],sep='')
# }
# save(myed.all,file='output/myed_all_05.RData')
# save(myed.mash.all,file = "output/myed_mash_all_05.RData")
load('output/myed_mash_all_05.RData')
out = summary_out(simdata$B,myed.mash.all)
knitr::kable(out$find_cond,col.names = name,caption = 'On finding conditions',digits = 5)
```

## Non-diag S

Try change $S_i$ to non-diagonal matrices. Generate $S$ from a wishart distribution with $\Lambda = I/20, df=10$.

```{r}
# generate S_i
# set.seed(12345)
# n = 1000
# S = rWishart(n,10,diag(5)/20)
# simdata = simple_sims01(500,S)
# data = mash_set_data(simdata$Bhat,simdata$Shat,V=S)
# strong = 1:nrow(simdata$B)
# U.c    = cov_canonical(data)
# U.pca  = cov_pca(data,5,strong)
# 
# myed.all = list()
# myed.mash.all = list()
# sigma2_list = c(0,1e-5,1e-4,1e-3,1e-2,1e-1,1)
# name = c()
# for(isig in 1:length(sigma2_list)){
#   myed.universal = myED_wrapper(data,U.pca,sigma2_list[isig],fix_sigma2 = TRUE,maxiter=100,tol=0.1)
#   myed.all[[isig]] = myed.universal
#   U.myed.universal  = myMash_data(myed.universal,data)$Ulist
#   m.c.myed.universal = mash(data, c(U.c,U.myed.universal),verbose = F,algorithm.version = 'R')
#   myed.mash.all[[isig]] = m.c.myed.universal
#   name[isig] = paste('sigma2=',sigma2_list[isig],sep='')
# }
# save(myed.all,file='output/myed_all_05_nondiagS.RData')
# save(myed.mash.all,file = "output/myed_mash_all_05_nondiagS.RData")

load("output/myed_mash_all_05_nondiagS.RData")
out = summary_out(simdata$B,myed.mash.all)
knitr::kable(out$find_cond,col.names = name,caption = 'On finding conditions',digits = 5)
```
