---
title: "Poisson log normal model simulation"
author: "DongyueXie"
date: "2020-07-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F,message = F)
```

## Introduction

Denote sample as $i=1,2,...,n$ and genes as $g =1,2,...,G$. Generate data from Poisson log Normal model \[Y_{ig}\sim Poisson(\text{e}^{\log(S_i*\frac{\exp({x_i^T\beta_g})}{\sum_g \exp({x_i^T\beta_g})})+\epsilon_{ig}-\frac{1}{2}\sigma^2_g}),\] where $\epsilon_{ig}\sim N(0,\sigma^2_g)$ and $S_i$ is the library size of sample $i$. The mean of $Y_{ig}$ is \[E(Y_{ig}) = S_i*\frac{\exp({x_i^T\beta_g})}{\sum_g \exp({x_i^T\beta_g})}\] and its variance is larger than its mean so overdispersion exists. 

The term $-\frac{1}{2}\sigma^2_g$ is only for illustration purpose and let the mean of $Y_{ig}$ interpretable. In practice, it will not affect the inference even if we generate data without this term, since we are comparing expressions given a gene, among samples. 

One advantage of the model is that we can induce dependence among genes in the latent space. For example, we have a dense $G\times G$ covariance matrix $\Sigma$ and we can generate a sample $Y_i$ of length $G$ in the following way: draw a length $G$ vector from \[Z_i\sim N(\log(S_i*\frac{\exp({x_i^T\beta_g})}{\sum_g \exp({x_i^T\beta_g})}),\Sigma)\] and draw \[Y_i|Z_i\sim Poisson(\exp(Z_i-\frac{1}{2}\text{diag}(\Sigma))\]. 

This simulation study compares (voom-limma, rle-voom-limma, deseq2)+(ashr). The number of genes is set to be $2000$, library size is $S_i = s_i*G$ where $s_i\sim Poisson(50)$ and number of simulations is 20.

Let's start with the simplest case where we have two conditions and each condition has equal number of samples. 


Assume there are 2000 genes and 3 replicates in each condition.

For the coefficients matrix $B\in R^{G*2}$, its first column $B_1$ is the intercept term and WLOG, are set to 1 and second column $B_2$ is the contrasts of two conditions. If all genes are null, then all elements of $B_2$ are 0; if some of genes(set to $10\%$ here) are non-null, then their $B_2$ are drawn from $N(0,0.8^2)$. 

## Diagnal Sigma all equal

```{r}
load("output/pln_simu_diagSigma01.RData")
source('code/pln_simu_summary.R')
summary_pln_simu(pln_simu)
```

## Diagnal Sigma from PLN fit

```{r}
load("output/pln_simu_diagSigma_plnFit.RData")
load("output/PLNfit_bulk_cytokin_SI.RData")
Sigma = diag(sigma(myPLN))
Sigma[which(Sigma>0.6)] = 0.3
hist(Sigma,breaks = 50)

summary_pln_simu(pln_simu)
```

## Full Sigma from PLN fit

```{r}
load("output/pln_simu_denseSigma_plnFit.RData")

summary_pln_simu(pln_simu)
```















